---
title: ' MRP model based on 2019 Canadian Federal Election'
author: "Xingyu Pu"
date: "2020/12/8"
output:
  html_document: default
  pdf_document: default
---
# 1. Abstract

One of the most important events in 2019, Canadian Federal Election, has ended. Justin Trudeau as leader of liberal party won 157 seats. However, there are lot of people who was able to vote did not vote in the election. Thus, this report is aiming to investigate is there would be much difference if we let everyone voted in the election by using multilevel regression and post-stratification method. Moreover, data in this report was supported by Canadian Election Study and General Social Survey.


## Keywords: 
2019 Canadian Federal Election, Multilevel Regression, Post-stratification, Canadian Election Study, General Social Survey



# 2. Introduction
The 2019 Canadian Federal Election has ended and the liberal party led by Prime Minister Justin Trudeau won 157 seats and formed minority government. In this report, it will be focused on exploring how would be different if everyone had voted in 2019 Canadian Federal Election. Let's review the result of 2019 Canadian Federal Election firstly.


party | votes | seats | percentage |
------|-------|-------|-----------
Liberal| 6,018,728 | 157 | 33.12 |
Conservative | 6,239,227 | 121  | 34.34 |
Bloc Quebecois | 1,387,030 | 32 | 7.63 |
New Democratic | 2,903,722 | 24 | 15.98 |
Green | 1,189,607 | 3  | 6.55 |
People's| 72,546 | 1 | 1.62 |



By calculating total number of voting people, we found there are only about 17 million people had voted in 2019 Canadian Federal Election. But the population is about 37 millions all over Canada according to 2019 to 2020 census of Statistics Canada. However, In Canada, people only over age 18 have right to vote. But not everyone over age 18 would like to vote in election. Therefore, it may bring some differences if we restrictively let everyone vote in the election. I will use multilevel regression post-stratification method to show the result in this report. Moreover, I'll use 2019 Canadian Election Study (CES) and General Social Survey (GSS) to illustrate how would be different if every person in Canada who was able to vote had voted in 2019 Canadian Federal Election.


This report include several parts. Firstly, I'll introduce and analyse survey data and census data. Then, from data I analyzed before, I can construct MRP models and interpret it. After obtaining MRP models, I can compare the estimated results from models with real results and then check if there are any differences. Lastly, I'll discuss weaknesses of my process.






```{r, echo=FALSE, message=FALSE}
survey_data <- read.csv('survey_data.csv')
census_data <- read.csv("census_data.csv")
```





# 3. Data Analysis
Nowadays, election is one of most basic elements of democracy. Canada has enhanced understanding of importance of democracy since 1965 and large amount of citizens conducted every election year. 2019 Canadian Election Study data included a dual-mode, two-wave data collection during the 2019 federal election and a post-election follow-up survey[3], which is a perfect survey sample. In the my survey data, I picked six variable which are gender, province, education level, vote for Trudeau or others and age groups. I choose them because they are more possible to bring differences on result of election. From CES data above, I can draw a histogram:
```{r, echo=FALSE, message=FALSE}
library(tidyverse)
plot1 = ggplot(survey_data, aes(Vote_Trudeau))+
  geom_bar(fill = 'yellow')
plot1
```

From the histogram, it is obvious that percentage of voting for Trudeau takes almost 50 percent of all voting accounts. Is that will be much difference if we let everyone had to vote then?



The Canadian General Social surveys provides data for academic research and teaching purposes only[4]. Thus, it is relatively reliable. My selected census data is General social survey on Canadians at Work and Home (cycle) in 2016. The target population concludes 4 more different human races and education level from 0 to 10. 


# 4. Model 
## logistics models
In order to see is there will be any differences is everyone voted in the election, I used logistics model to estimate it and compare it with actual result.

The two logistic models below are affected by two aspects :gender and education level. Each of them has different effect on rate of voting.

**gender:** Since voters have difference genders and sometimes it may be biased. For example, woman may prefer to vote for parties which lead by woman.

**education level:** Educational level is important on accounting voting rates. Since voters with different educational levels may have different voting preference. For example, voters who have higher educational level usually prefer to vote on parties which have some scientific accomplishments.



```{r, echo=FALSE, include=FALSE, message=FALSE}
# Creating logistic Models for both Trudeau and Others results
model_Trudeau <- glm(Vote_Trudeau ~ gender  + edu_level , 
             data=survey_data, family= "binomial")

model_others <- glm(Vote_others ~ gender  + edu_level , 
             data=survey_data, family= "binomial")
```


```{r,echo=FALSE, include=FALSE, message=FALSE}
model_Trudeau
```

**Model 1**
$$\log{(\frac{\hat{p_1}}{1-\hat{p_1}})} =-1.69215-0.03491X_{genderA woman}-0.33667X_{genderOther}+0.12099X_{edu_level}$$

**Interpretation:** $\hat{\beta_0}$ = -1.69215 represents the intercept of the model, $\hat{\beta_1}$ = -0.03491, which shows effect to vote for Trudeau when gender is female. 
$\hat{\beta_2}$ = -0.33667, which shows effect to vote for Trudeau when gender is not female.
$\hat{\beta_3}$ =  0.12099, which shows effect of education level on voting for Trudeau.

From this model, it is clear that woman have negative effect on voting for Trudeau, same as others gender. However, voters with others gender have more negative effect on voting for Trudeau comparing to female voters. In contrast, educational level is proportional to voting for Trudeau, which has positive effect on voting for Trudeau.

               



```{r,echo=FALSE, include=FALSE, message=FALSE}
model_others
```


**Model 2**
$$\log{(\frac{\hat{p_2}}{1-\hat{p_2}})} =1.69215+0.03491X_{genderA woman}+0.33667 X_{genderOther} -0.12099X_{edu_level}$$

**Interpretation:** $\hat{\beta_0}$ = 1.69215 represents the intercept of the model, $\hat{\beta_1}$ = 0.03491, which shows effect to vote for others when gender is female. 
$\hat{\beta_2}$ = 0.33667, which shows effect to vote for others when gender is not female.
$\hat{\beta_3}$ = -0.12099, which shows effect of education level on voting for others.

In this model, from coefficients above, it is clear to show that woman have positive effect on voting for Other parties, which is proportional to it, same as others gender. However, voters with others gender have larger effect on voting for other parties comparing to female voters. In contrast, educational level is inversely proportional to voting for other parties, which has negative effect on it.

It is obvious that trend and coefficients of model 1 and model 2 are opposite. 





## post-stratification

Moreover, in order to estimate people's preference on voting party in large population, I 'll use post-stratification method. 
Firstly, I divided original data into different groups based on races, gender and education level. Then, I can calculate logistics estimator:
$\hat{y_{ij}}$ = $$\log{(\frac{\hat{p_{ij}}}{1-\hat{p_{ij}}})}=\hat{\beta_0}+\hat{\beta_1}X_{genderA woman_{ij}}+\hat{\beta_2}X_{genderOther_{ij}}+\hat{\beta_3}X_{edu_level_{ij}}$$

Then, after doing some mathematics calculation, I can get:
$$  \hat{p_{ij}} ={\frac{e^{\hat{y_{ij}}}}{e^{\hat{y_{ij}}}+1}} $$

Finally, the post-stratification outcome will be:
$$  \hat{p_i}^{ps} = \frac{\sum{N_j}\hat{p_{ij}}}{\sum{N_j}} $$
Here ${N_j}$ is the number of each group. $\hat{p_{ij}}$ shows estimated probability of voters who are in group j have preference on party i. $\hat{p_i}^{ps}$ indicates estimated probability of voters who have preference to vote for Trudeau or others. 




# 5. Results


From logistics models and post-stratification method above, we can calculated the estimated probability of voters for voting each party. The results is shown below:

party | votes percentage | 
------|-------|
Liberal| 28.68 |
Conservative | 60.57 |


In the result, it is clear to see liberal party led by Trudeau still takes large proportion of total voting. Although the estimated result is not totally same with actual result, the difference is not huge, so it can be considered as a successful estimated result. However, we noticed that total sum of percentage is less than 100. This is because in survey data we selected, there a lot of voters chose 'Don't know/prefer not to answer'. Therefore, the estimated result looks smaller than actual one.







# 6. Discussion

In this report, I chose 2019 Canadian Election Study (CES) as survey sample and General Social Survey (GSS) in 2016 as a census sample. The reason I chose them is they are unbiased and relative accurate. Therefore, the result based on the data I selected is reliable. Now, let's go back to the very first question I raised in beginning: Is that will be different if we let everyone who was able to vote voted in the election? From result in last part, the answer is yes, but not much since the difference is little. Therefore, the overall ratio for voting on Trudeau is not changed.



# 7.Weakness

Although there is no big difference between my final estimated result with actual result, there are still some weaknesses in my process:
In my report, I divided all Canadian parties into only 2 parts: Liberal party and others, which is too general. Moreover, I only focused on differences of voting for Trudeau, so it may cause over-generalization. In addition, The survey data and census data I selected are not latest. Survey data is CES in 2019 and census data is GSS in 2016. Therefore, the data maybe not accurate enough.

## Next steps
In next time, I will analyze each party deeply and check their differences after everyone has voted instead of only focusing on differences of voting for Trudeau. Moreover, I will use more method and technique to construct models in order to get better estimation and select newest census data to obtain more precise data.



# Appendix

A link to github repository: https://github.com/xingyupu/Final-project


# 8. References citation


1. 2019 Canadian federal election. (2020, December 03). Retrieved December 09, 2020, from https://en.wikipedia.org/wiki/2019_Canadian_federal_election


2. Statistics Canada. Table 17-10-0009-01 Population estimates, quarterly.


3. Stephenson, Laura B; Harell, Allison; Rubenson, Daniel; Loewen, Peter John, 2020, "2019 Canadian Election Study - Online Survey", https://doi.org/10.7910/DVN/DUS88V, Harvard Dataverse, V1


4. General social survey on Canadians at Work and Home (cycle 30), 2016. (n.d.). Retrieved December 23, 2020, from https://sda-artsci-utoronto-ca.myaccess.library.utoronto.ca/sdaweb/html/gss.htm





